Supporting Information: Unsupervised embedding of trajectories captures the latent structure of mobility

### 2022-11-28 paper/current/main.tex
second_round_revision

Armed with this result, we can now show the equivalence between the SGNS {\it word2vec} model and the gravity model.
We set the window length to $w=1$ to restrict {\it word2vec} to predict the first-order flux $\hat T_{ij}$ between locations, as is the case for the gravity model.
Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimal in-vectors and out-vectors such that $\vect{v}_i = \vect{u}_i$ \autocite{levy2014neural}.
By setting $\gamma=1$,  we have 
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} \propto P(i) P^{NS}(j\given i) \propto \frac{P(i)P(j)\exp(\bm{u}_j \cdot \bm{u}_{i})}{Z_i}.
\end{align}
The flow $\widehat{T}_{ij}$ is symmetric (\ie $\widehat{T}_{ij}=\widehat{T}_{ji}$) because the skip-gram model neglects whether the context $j$ appears before or after the target $i$ in the trajectory.
If we swap $i$ and $j$ in Eq.~\eqref{eq:flow_w2v}, the numerator remains the same but the denominator can be different.
Therefore, to ensure $\widehat{T}_{ij}=\widehat{T}_{ji}$, the denominator $Z_i$ should be a constant.
Taken together, the {\it word2vec} model with the negative sampling predicts a flow in the same form as the gravity model:
\begin{align}
	\hat{T}_{ij}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}). \label{eq:flow_w2v_ng0}
\end{align}

@@

@@

We report that this curve is almost universal across many countries.
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset).
Other nations exhibit different curves which lack the portions with decreasing norm, probably indicating the lack of internationally-prestigious institutions.
Similar patterns can be found across many of the 30 countries with the most total researchers (Fig.~\ref{fig:concave30}; see \SI{} for more discussion).

@@

While we focus on three domains of human mobility and illustrate its potential using scientific mobility, this approach is general and can be applied to many different domains, such as animal migration, transit-network mobility, and international trade.
Once learned, functional distances between locations, such as countries, cities, or organizations, or the embedding model itself, can be published to facilitate re-use, and support reproducibility and transparency in cases when the underlying data is too sensitive to be made available.
Moreover, this approach can be used to learn a functional distance even between entities for which no geographic analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, we also demonstrate how the topology of the embedding space can be leveraged to facilitate interpretation and application of the complex features of mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Other operations such as comparing entities or calculating aggregates, which could be complex and computationally expensive for other methods, are here reduced to simple vector arithmetic.
Embeddings also allow us to quantitatively explore abstract relationships between locations, such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

>>>>

Armed with this result, we can now show the equivalence between the SGNS {\it word2vec} model and the gravity model.
We set the window length to $w=1$ to restrict {\it word2vec} to predict the first-order flux $\hat T_{ij}$ between locations, as is the case for the gravity model.
Parameter $\gamma=1$ is a special choice that ensures that, when the embedding dimension is sufficiently large, there exists optimal in-vectors and out-vectors such that $\vect{v}_i = \vect{u}_i$ \autocite{levy2014neural}.
By setting $\gamma=1$,  we have 
\begin{align}
	\label{eq:flow_w2v}
	\hat{T}_{ij} \propto P(i) P^{NS}(j\given i) \propto \frac{P(i)P(j)\exp(\bm{u}_j \cdot \bm{u}_{i})}{Z_i}.
\end{align}
The flow $\widehat{T}_{ij}$ is symmetric (\ie $\widehat{T}_{ij}=\widehat{T}_{ji}$) because the skip-gram model neglects whether the context $j$ appears before or after the target $i$ in the trajectory which produces,
\begin{align}
	\label{eq:unbiasef2}
	T_{ij}=T_{ji} &\iff \frac{P(j) f(\vect{u}_j \cdot \vect{u}_{i}) }{Z_i} P(i) = \frac{P(i)  f(\vect{u}_i \cdot \vect{u}_{j}) }{Z_j} P(j) \nonumber \\
	&\iff \frac{1}{Z_i}= \frac{1}{Z_j}  \nonumber \\
	&\iff Z_i = Z_j
\end{align}
Taken together, the {\it word2vec} model with the negative sampling predicts a flow in the same form as the gravity model:
\begin{align}
	\hat{T}_{ij}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}). \label{eq:flow_w2v_ng0}
\end{align}

@@

We further demonstrate the ability of \textit{word2vec} to effectively represent gravity-like relationships with an synthetic benchmark. Namely, we train a \textit{word2vec} model using synthetic mobility trajectories that strictly adhere to the gravity model (see Supporting Information for details). 
We find that distances in the embedding space strongly correlates with distances in the synthetic space which was explicitly structured according to the gravity model (Pearson correlation of 0.943 and 0.801, depending on the distance metric used, Fig. S4).
The results of this experiment align with a previous study which interpreted \textit{word2vec} as able to recover the metric underlying a semantic manifold~\autocite{hashimoto2016word}.

@@

We report that this curve is almost universal across many countries.
For instance, China's curve closely mirrors that of the United States (Fig.~\ref{fig:length}b).
Smaller but scientifically advanced countries such as Australia and other populous countries such as Brazil also exhibit curves similar to the United States (Fig.~\ref{fig:length}b, inset).
Other nations exhibit different curves which lack the portions with decreasing norm, probably indicating the lack of internationally-prestigious institutions.
Similar patterns can be found across many of the 30 countries with the most total researchers (Fig.~\ref{fig:concave30}; see \SI{} for more discussion).  Although we speculate on potential explanations for this pattern (see \SI{} for more discussion), we don't intend to make general claims about scientific mobility, or mobility more generally, but rather to illustrate how \textit{word2vec} can facilitate new kinds of investigations into mobility.
Further work is necessary to determine what causes this curve pattern in so many countries and whether the same pattern can be found in other domains of human mobility.

@@

While we focus on three domains of human mobility and illustrate its potential using scientific mobility, this approach is general and can be applied to many different domains, such as animal migration, transit-network mobility,  discretized cell-phone location data,  and international trade.
Once learned, functional distances between locations, such as countries, cities, or organizations, or the embedding model itself, can be published to facilitate re-use, and support reproducibility and transparency in cases when the underlying data is too sensitive to be made available.
Moreover, this approach can be used to learn a functional distance even between entities for which no geographic analog exists, such as between occupational categories based on individuals' career trajectories.
In addition to providing a functional distance that supports modeling and predicting mobility patterns, we also demonstrate how the topology of the embedding space can be leveraged to facilitate interpretation and application of the complex features of mobility.
As we have shown, the embedding space allows the visualization of the complex structure of scientific mobility at high resolution across multiple scales, providing a large and detailed map of the landscape of global scientific mobility.
Other operations such as comparing entities or calculating aggregates, which could be complex and computationally expensive for other methods, are here reduced to simple vector arithmetic.
Embeddings also allow us to quantitatively explore abstract relationships between locations, such as academic prestige, and can potentially be generalized to other abstract axes.
Investigation of the structure of the embedding space, such as the vector norm, reveals universal patterns based on the organization's size and their vector norm that could be leveraged in future studies of mobility.

### 2022-11-28 paper/current/supp.tex
second_round_revision

In the case that an individual lists multiple affiliations in a single year, affiliations listed on publications published in that year are shuffled between each iteration of the \textit{word2vec} training process (each epoch).
For example, an author who published with affiliation $A$ in $t_{0}$, and affiliations $B$ and $C$ in $t_{1}$ could appear as one of $(A, B, C)$ or $(A, C, B)$ in each training iteration.
This effectively removes the effect of order within a year, as the order cannot be meaningfully established based on co-affiliations in a single paper, or on different affiliations listed on separate papers, for which the date of publication may not be representative of the actual completion of the project.

@@

%
% S4 TEXT
\paragraph*{S4 Text}
\label{si:text:ppr_dist}
{\bf Network-based personalized Page Rank distances.}

@@

%
% S5 Text
\paragraph*{S5 Text}
\label{si:text:Raw SVD}
{\bf Singular value decomposition distance.}

@@

%
% S6 Text
\paragraph*{S6 Text}
\label{si:text:Laplacian}
{\bf Laplacian Eigenmap distance.}

@@

%
% S8 Text
\paragraph*{S8 Text}
\label{si:text:direct}
{\bf Direct optimization of gravity model}

@@

%
% S9 Text
\paragraph*{S9 Text}
\label{si:text:whynueralbetter}
{\bf Why do neural embeddings outperform direct matrix factorization?}

@@

%
% S10 TEXT
\paragraph*{S10 Text}
\label{si:text:organizations}
{\bf Organization disambiguation and metadata.}

@@

%
% S11 TEXT
\paragraph*{S11 Text}
\label{si:text:disambiguation}
{\bf Author name disambiguation.}

@@

% S12 TEXT
\paragraph*{S12 Text}
\label{si:text:disambiguation}
{\bf Derivation of Eq.\ref{eq:nce} - Noise Contrastive Estimation}

@@

%
% S13 TEXT
\paragraph*{S13 Text}
\label{si:text:network_ranking}
{\bf Reconstructing Times ranking with network measure.}

@@

%
% S14 TEXT
\paragraph*{S14 Text}
\label{si:text:boomerang_speculation}
{\bf Speculation on variations of the convex-curve pattern.}

@@

%
% S15 TEXT
\paragraph*{S15 Text}
\label{si:text:generalized_gravity_model}
{\bf Generalized gravity model.}

@@

\multicolumn{1}{!{\huxvb{0, 0, 0}{0}}l!{\huxvb{0, 0, 0}{0.5}}}{\huxtpad{6pt + 1em}\raggedright \hspace{6pt} \textbf{PPR JSD} \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.888 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.847 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0.5}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.835 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.871 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.828 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0.5}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.814 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.865 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.847 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.834 \hspace{6pt}\huxbpad{6pt}} \tabularnewline[-0.5pt]

@@

>>>>

In the case that an individual lists multiple affiliations in a single year, affiliations listed on publications published in that year are shuffled between each iteration of the \textit{word2vec} training process (each epoch).
For example, an author who published with affiliation $A$ in $t_{0}$, and affiliations $B$ and $C$ in $t_{1}$ could appear as one of $(A, B, C)$ or $(A, C, B)$ in each training iteration.
This effectively removes the effect of order within a year, as the order cannot be mFigeaningfully established based on co-affiliations in a single paper, or on different affiliations listed on separate papers, for which the date of publication may not be representative of the actual completion of the project.

@@

\paragraph*{S4 Text}
\label{si:text:synthetic_exp}
{\bf Demonstration on ability of \textit{word2vec} to effectively represent gravity-like relationships with a simulation experiment}

We conducted an additional experiment to test the ability of \textit{word2vec} to recover the metric between locations, and to further demonstrate its equivalency with the gravity model. Specifically, we perform a synthetic benchmark to test whether \textit{word2vec} can effectively recover gravity-like relationships in data that is known to contain them.

We start by generating synthetic mobility trajectories structured according the gravity model. For this purpose, we first generate an artificial ``geographic'' space. 1,000 points are generated in a 2-dimensional space, with their coordinates distributed evenly between -1 and 1. These points act as the ``locations'' in our simulation. Each point is also assigned a ``mass'' randomly drawn from a power-law distribution $p(x) \propto x^{-k}$ (we use $k=2$) that is meant to represent the ``population'' or ``size'' of each location.

With this simulated geography in place, we calculate the pairwise distance matrix $r_{ij}^{synthetic}$, which is then used to calculate a transition matrix between points, defined as $T_{ij} = M_i M_j f(r_{ij})$, with transition probabilities following the gravity law. The purpose of this transition matrix is to define the likelihood of movement between points (or \textit{locations}) in our simulated geography based on their mass and pairwise distances. We conduct this experiment with two different transition matrices, one based on cosine distance where $f(r_{ij})=e^{-\beta r_{ij}}$ and another with euclidean distance where $f(r_{ij})=r_{ij}^{-\alpha}$. For simplicity, we use $\beta=2$ and $\alpha=2$, though the analysis was repeated for multiple parameters in order to assess robustness.

Finally, synthetic trajectories are generated based on the probabilities in the transition matrix.
Specifically, we use a random walker based approach, where random walkers are initialized in locations with a probability proportional to their mass and transition between points according to the transition matrix. 
These synthetic trajectories are then used to train the skip-gram \textit{word2vec} model. 
For parameters, we set window size to $w=1$ and the embedding dimension to $d=100$.

From the embedding trained with synthetic trajectories, we calculate the pairwise embedding distances between points, $r_{ij}^{embedding}$.
To evaluate the ability \textit{word2vec} to recover the known gravity-like structure of the synthetic data, we compare the embedding distance to the distances between the 1,000 points in the synthetic space, $r_{ij}^{synthetic}$ using the Pearson correlation coefficient. 
This result is repeated 100 times, and the correlation coefficients are aggregated across iterations.

We find that \textit{word2vec} effectively recovers the gravity-like relationships in the synthetic data. 
Based on cosine distance, the Pearson correlation between the embedding and synthetic distances is 0.943, with a standard error of 0.01. 
Based on euclidean distance, the correlation was 0.801, with a standard error of 0.01.
To better illustrate this point, Fig. \ref{fig:supp:synthetic_exp} shows the correlation for one representative iteration of this experiment.

\

%
% S4 TEXT
\paragraph*{S5 Text}
\label{si:text:ppr_dist}
{\bf Network-based personalized Page Rank distances.}

@@

%
% S5 Text
\paragraph*{S6 Text}
\label{si:text:Raw SVD}
{\bf Singular value decomposition distance.}

@@

%
% S6 Text
\paragraph*{S7 Text}
\label{si:text:Laplacian}
{\bf Laplacian Eigenmap distance.}

@@

%
% S8 Text
\paragraph*{S9 Text}
\label{si:text:direct}
{\bf Direct optimization of gravity model}

@@

%
% S9 Text
\paragraph*{S10 Text}
\label{si:text:whynueralbetter}
{\bf Why do neural embeddings outperform direct matrix factorization?}

@@

%
% S10 TEXT
\paragraph*{S11 Text}
\label{si:text:organizations}
{\bf Organization disambiguation and metadata.}

@@

%
% S11 TEXT
\paragraph*{S12 Text}
\label{si:text:disambiguation}
{\bf Author name disambiguation.}

@@

% S12 TEXT
\paragraph*{S13 Text}
\label{si:text:disambiguation}
{\bf Derivation of Eq.\ref{eq:nce} - Noise Contrastive Estimation}

@@

%
% S13 TEXT
\paragraph*{S14 Text}
\label{si:text:network_ranking}
{\bf Reconstructing Times ranking with network measure.}

@@

%
% S14 TEXT
\paragraph*{S15 Text}
\label{si:text:boomerang_speculation}
{\bf Speculation on variations of the convex-curve pattern.}

@@

%
% S15 TEXT
\paragraph*{S16 Text}
\label{si:text:generalized_gravity_model}
{\bf Generalized gravity model.}

@@

\multicolumn{1}{!{\huxvb{0, 0, 0}{0}}l!{\huxvb{0, 0, 0}{0.5}}}{\huxtpad{6pt + 1em}\raggedright \hspace{6pt} \textbf{PPR JSD} \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.888 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 
7 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0.5}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.835 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.871 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.828 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0.5}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.814 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.865 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.847 \hspace{6pt}\huxbpad{6pt}} &
\multicolumn{1}{r!{\huxvb{0, 0, 0}{0}}}{\huxtpad{6pt + 1em}\raggedleft \hspace{6pt} 0.834 \hspace{6pt}\huxbpad{6pt}} \tabularnewline[-0.5pt]

@@

\newpage
%
% Figure - reverse CDF by scale
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=0.8\textwidth]{\figdir/rebuttal/test_result.pdf}
	    \caption{\textbf{Neural embedding recovers the underlying distance from trajectories. }\textbf{a.} The trajectories generated with cosine distance an \textbf{b.} The trajectories generated with euclidean distance. Illustratively, we show the result of the representative iteration of the experiment whose correlation is close to the average. The embedding distance shows a good agreement with underlying distance. Orange dots represent the average embedding distance across binned synthetic distances with standard error.
    }
	\label{fig:supp:synthetic_exp}
\end{figure}
